\chapter{Introduction}
\label{chapter:introduction}

\section{Motivation}

Shared nothing architecture%
\sidenote[5]{
  \emph{Shared Nothing}: A type of architecture where the servers do not share
  any hardware resources between them.
} \citep{Stonebraker86g} is a popular distributed system design used for
building high traffic web applications. The servers within such a system setup
sometimes need to synchronize their activities or need to access a shared
resource. A service used for such a purpose could act as a single point of
failure based on its design and the system architecture.

Online social game Magic Land uses a stateful application server architecture to
handle millions of users. This architecture uses Redis as a global registry to
keep track of user sessions. Failure of the software Redis, or the server it
is deployed on can lead to application downtime. This creates a need for
building a service that is itself distributed, fault tolerant and primarily
consistent.

\section{Goals}

The goal of this thesis is to build a customized locking service focussed on
being fault tolerant without compromising on consistency and performance.
Specifically,

\begin{itemize}
    \iterm{Strong consistency}: The state of the system remains valid across
    all concurrent requests.
    \iterm{High availability}: The system should be operable at all times.
    \iterm{Performance}: The system should be able to handle large number of
    requests.
    \iterm{Fault tolerant}: Server failures should be isolated and the system
    should continue to function despite these failures.
\end{itemize}

We investigate the algorithms for distributed consensus and analyze similar
projects. We then create a system design based on these observations and
an working implementation.

\section{Scope}

The scope of the project is to implement a reasonably stable application that
satisfies the above goals and can be used in production. The application needs
to have good test coverage and documented code. The thesis scope does not cover
creation of any new algorithms or techniques, but rather builds on the basis of
well known ideas in the field.

\section{Overview}

Consensus in distributed systems is a large topic in itself. They are
notoriously hard to implement even though the pseudo code describing them can be
relatively simple. Among the set of algorithms avaiable for arriving at
consensus in an asynchronous distributed system, we choose Paxos because of its
literature coverage and usage in the industry.

The Erlang programming language is built with the primitives necessary for
creating a distributed system, such as, stable messaging, crash tolerance and
well defined behaviours. The Actor model of Erlang processes maps very well to 
the algorithms pseudo code. Using Erlang allows us to separate the details of
the algorithm from the imlementation specifics. This enables us to ensure
accuracy and makes debugging simpler.

The idea behind the system design is that we use Multi-Paxos for consensus,
reconfigurable state machine algorithms to make the cluster dynamic and
implement this in Erlang.

\section{Organization}
The thesis is divided into three main sections over multiple chapters.

The first section provides the context and background information related to the
project. It details the problem background, related projects, the research
area and the set of requirements the project is based on.

The second section describes the analysis, design, implementation of the project
and the experiments performed on it.

The final section discusses the project results and provides an outline for
the future work.

