\chapter{Preface}

The idea behind Warlock was born of a real project need at Wooga by Paolo Negri
and Knut Nesheim. Paolo, as the thesis supervisor, helped me out with the
project's conceptualization, feature set and figuring out some of the hard
problems during the course of the project. Knut was very patient and helpful
when trying to push Erlang to its limits. My deepest gratitude to them both.

Justin Pearson, Senior Lecturer at the Department of Information Technology,
Uppsala University gracefully accepted to be the thesis reviewer and provided
valuable input for the project. Olle Eriksson, Senior Lecturer at the Department
of Information Technology, Uppsala University was really helpful as thesis
coordinator with his excellent organization.

I would like to thank Anders Berglund, Ivan Christoff and staff of
of the Information Technology Department for taking care of all the
administrative formalities.

Wooga GmbH, Berlin hosted me for five months and provided me with all the
resources for my thesis.

This project would not have been possible without the enthusiastic support from
the people at Wooga, the Magic Land team and the Wooga back end team. Thanks to
Marc Hämmerle for the lively project discussions and the name \term{Warlock}.

I would also like to thank Olle Gällmo from Uppsala University for the
Project \abbr{CS} course. It allowed us to work on a large Erlang project and
visit the Erlang User Conference 2011, Stockholm, where I was
introduced to Wooga.

Thanks to Eivind Uggedal for compiling this beautiful {\LaTeX} template.

Lastly, I would like to thank the scientific and open
source community for all their contributions to further the progress in the
field of distributed computing.\\

{\raggedleft{\scshape Sukumar Yethadka}\par}

{\raggedright{Uppsala, Sweden \\ September, 2012}\par}

